Realizing Temporal Graphs

### 2022-11-29 RealizingTemporalGraphs.tex
Tuesday 29th November - Overleaf changes

@@

\problemdef{\kDeltaUpperBoundLong}
{An $n \times n$ matrix $D$ of positive integers.}
{Does there exist a graph $G=(V,E)$ with vertices $\{v_1,\ldots,v_{n}\}$ 
and a $\Delta$-periodic labeling $\lambda: E \rightarrow \{1,2,\ldots,\Delta\}^k$ such that, 
for every $i,j$, the duration of the fastest temporal path from $v_i$ to $v_j$ in the $\Delta$-periodic temporal graph $(G,\lambda,\Delta)$ is \emph{at most} $D_{i,j}$.}

\problemdef{\kDeltaLowerBoundLong}
{An $n \times n$ matrix $D$ of positive integers.}
{Does there exist a graph $G=(V,E)$ with vertices $\{v_1,\ldots,v_{n}\}$ 
and a $\Delta$-periodic labeling $\lambda: E \rightarrow \{1,2,\ldots,\Delta\}^k$ such that, 
for every $i,j$, the duration of the fastest temporal path from $v_i$ to $v_j$ in the $\Delta$-periodic temporal graph $(G,\lambda,\Delta)$ is \emph{at least} $D_{i,j}$.}

@@

\section{
\texorpdfstring{\deltaExact\ is polynomial-time solvable}
{\textsc{Simple Delta-TGR} is polynomial-time solvable}
}\

    \subsection*{Cycles}
    \begin{definition}[Travel delays]
    Let $e_1=uv$ and $e_2=vz$ be two incident edges with $e_1 \cap e_2 = v$.
    We define the \emph{travel delay} from $u$ to $z$ at vertex $v$, denoted with $\tau_v^{uz}$,
    as the difference of the labels of $e_2$ and $e_1$, where we subtract the value of the label of $e_1$ from the label of $e_2$,  modulo~$\Delta$.
    More specifically:
    \begin{equation}\label{eq:def-VertexWaitingTime}
       \tau_v^{uz} \equiv \lambda (e_2) - \lambda(e_1) \pmod \Delta.
    \end{equation}
    Similarly, $\tau_v^{zu} \equiv \lambda (e_1) - \lambda(e_2) \pmod \Delta$.
    \end{definition}
    Intuitively, the value of $\tau_v^{uz}$ represents how long a temporal path waits at vertex $v$ when first taking the edge $e_1=uv$ and then the edge $e_2 = vz$.

@@

    \begin{observation} \label{obs:exact-VertexSum}
    Let $v$ be a vertex of $G$, and let $N(v)=\{u_1, u_2, \ldots, u_k\}$, Then 
    \begin{equation*}
        \sum_{i = 1}^k \tau_{v}^{u_{i}v_{i+1}} \equiv  0 \pmod \Delta,
    \end{equation*}
    where the indices of vertices $u_i$ are taken modulo $k$.
    \end{observation}

    \begin{proof}
        Using \cref{eq:def-VertexWaitingTime} and the fact that the indices of vertices $v_i$ are taken modulo $k$, we get the following:
        \begin{align*}
            &\sum_{i = 1}^k \tau_{v_i}^{v_{i-1}v_{i+1}}  \equiv  
            \sum_{i = 1}^k  ( \lambda(v_i v_{i+1}) - \lambda (v_{i-1} v_i) ) \pmod \Delta \equiv \\
            &\lambda(v_1 v_2) - \lambda(v_{k}v_1) + \lambda(v_2 v_3) - \lambda(v_2 v_1) + \dots + \lambda(v_k v_1) - \lambda(v_{k-1}v_k) \pmod \Delta.
        \end{align*}
    \end{proof}

\subsection{
\texorpdfstring{\deltaExact\ } {Simple Delta-TGR}
is polynomial-time solvable on trees, paths and cycles}

Let us observe some properties of the matrix $D$ from \deltaExact, when the underlying graph $G$ of $D$ is a cycle $C_n = \{v_1, v_2, \dots, v_n\}$  on $n$ vertices.
By the definition, each vertex is on distance $0$ from itself. This corresponds with all diagonal entries of $D$ being $0$.
Now, let us observe that each vertex $v_i$ has exactly two neighbours in $C_n$, namely $v{i-1}$ and $v_{i+1}$,
therefore for all $i\in [n]$ we set $D_{i,i-1} = D_{i, i+1} = 1$, where indices are taken modulo $n$.
This results in the upper and lower diagonal of $D$ having all $1$'s, with $D_{1,n} = D_{n,1} = 1$.
The matrix $D$ is of the following form
\begin{equation*}
  D = 
  \begin{bmatrix}
    0 & 1 &  &  &  & 1\\
    1 & 0 & 1 &  &  & \\
    & 1 & 0 & 1 & & \\
    & & \ddots & \ddots & \ddots &   \\
    &   & &1 &  0 & 1\\ 
    1 &  &  &  &1 &  0 
  \end{bmatrix},
\end{equation*}
where the empty entries consists of positive integers different than $1$.

Let $v_i$ be an arbitrary vertex in $C_n$.
Vertex $v_i$ can reach an arbitrary vertex $v_k \in C_n$ using a positive side of the cycle (\ie going from $v_i$ to $v_{i+1},  v_{i+2}$, etc.), 
and using the negative side of the cycle (\ie going from $v_i$ to $v_{i-1}, v_{i-2}$, etc.).
Let us denote with the $d^+(i,k)$ the duration of the temporal path from $v_i$ to $v_k$ using the positive side of the cycle,
and with $d^-(i,k)$, the duration of the temporal path from $v_i$ to $v_k$ using the negative side of the cycle.
Since these two are the only possible paths from $v_i$ to $v_k$ in $C_n$ we know that $D_{i,k} = min \{d^+(i,k), d^-(i,k)\}$.

\begin{claim} \label{claim:poly-delta-cycle-monotonicity}
    If vertex $v_i$ reaches vertex $v_j$ the fastest, using the positive (resp.~negative) side of the cycle, \ie $v_i, v_{i+1}, \dots, v_{j-1}, v_j$ (resp. $v_i, v_{i-1}, \dots, v_{j+1}, v_j$),
    then $v_i$ reaches all other vertices $v_k$, where $k \in \{i+1, i+2, \dots, j-1\}$ (resp.~ $k \in \{i-1, i-2, \dots, j+1\}$)
    using the same, positive side  (resp.~negative), side of the cycle,
    where the indices are taken modulo $n$.
\end{claim}
\begin{proof}
    Let $v_i, v_j$ be two arbitrary vertices in $C_n$ and suppose that $D_{i,j} = d^+(i,j)$,
    but there exists a vertex $v_k$ from $C_n$, where $k \in \{i+1, i+2, \dots, j-1\}$,
    for which $D_{i,k} \neq d^+(i,k)$.
    Then $d^-(i,k)$ passes through the vertex $v_j$, and we can split the path from $v_i$ to $v_k$ into two pieces, 
    first from $v_i$ to $v_j$ and the second one from $v_k$ to $v_j$.
    So 
    \begin{equation} \label{eq:poly-delta-cycle-monotonicity}
        d^-(i,k) = d^-(i,j) + d^-(j,k)^*,
    \end{equation} where $d^-(j,k)^*$ is the duration of the path from $v_j$ to $v_k$, using the negative side of the cycle, with consideration that we come to the vertex $v_j$ at time $d^-(i,j)$ and we potentially have to wait there for some positive amount of time, before we continue to $v_j$
    (\ie $d^-(j,k)^* \geq d^-(i,j) \geq D_{j,k}$ includes some waiting time at vertex $v_j$).
    By the assumption we know that $d^+(i,j) \leq d^-(i,j)$, so we can rewrite \cref{eq:poly-delta-cycle-monotonicity} as
    $d^-(i,k) \leq d^+(i,j) + d^-(j,k)^*$. This means that we can reach $v_k$ from $v_j$ by going from $v_i$ to $v_j$ using the positive side of the cycle, wait some time at $v_j$, before we continue back to $v_j$.
    In the above construction vertex $v_k$ is visited twice. By the definition, the fastest temporal path from $v_i$ to $v_k$ visits $v_k$ exactly once. Therefore we can stop at $v_k$ already when traveling from $v_i$ to $v_j$ the first time, using the positive side of the cycle.
    It follows that $D_{i,j} = d^+(i,j) \leq d^-(i,j)$, which contradicts with our assumption.
\end{proof}

Let $v_i$ be an arbitrary vertex of $C_n$. Let us look at the row $i$ of the matrix $D$, which corresponds to the duration of fastest temporal paths from $v_i$ to any other vertex $v_j \in C_n$.
Using \cref{claim:poly-delta-cycle-monotonicity} we know that $v_i$ will reach some consecutive vertices $v_{i+1}, v_{i+2}, \dots, v_j$ the fastest, using the positive side of the cycle 
and $v_{i-1}, v_{i-2}, \dots, v_{j+1}$ the fastest, using the negative side of the cycle.
Suppose $v_j \in C_n$ is the last vertex $v_i$ reaches using the positive side of the cycle, and $v_{j+1}$ the last vertex that is reached using the negative side of the cycle.
Then we know that $D_{i,i+1} < D_{i,i+2} < \cdots< D_{i,j-1} < D_{i,j}$ and 
$D_{i,i-1} < D_{i,i-2} < \cdots < D_{i,j+2} < D_{i,j+1}$.
Note that it can happen that $v_j = v_{j+1}$, \ie to reach vertex $v_j$ from $v_i$ the fastest, we can use either positive or negative side of the cycle.
Using the above observations, every row $i$ ($i \in [n]$) of matrix $D$ has two (or one) maximum elements, one at position $j$ and the other at position $j+1$, where $j, j+1 \in [n]$ and the indices are considered modulo $n$.
Let us denote these two values as $m_i^1$ and $m_i^2$.
The row $i$ of $D$ is of the following form, it has a $0$ at the entry $i$, it has a $1$ at entries $i-1, i+1$,
the values increase on the positions $i+1, i+2, i+3, \dots, j-1, j$ for some $j$ with value $m_i^1$, 
and on the other side, values increase on the positions $i-1, i-2, i-3, \dots, j+2, j+1$ for some $j+1$ with value $m_i^2$,
where indices are taken modulo $n$.

\begin{theorem}
    \deltaExact\ can be solved in polynomial time on cycles.
\end{theorem}

>>>>

\section{Ideas}
\begin{itemize}
    \item Symmetric distance matrix $D$: polytime?
    \item Use $|D|_\infty$ (maximum value in $D$) + max degree $\Delta$ as parameter $\rightarrow$ should give FPT
    \item FPT with feedback edge number
    \item use as parameter $\max_{i,j} |D_{i,j}-D_{j,i}|$
    \item approximation (additive or multiplicative), maybe with graph as input
\end{itemize}

@@

\problemdef{\kDeltaUpperBoundLong}
{An $n \times n$ matrix $D$ of positive integers and an integer $k\in \mathbb{N}$.}
{Does there exist a graph $G=(V,E)$ with vertices $\{v_1,\ldots,v_{n}\}$ with $n+k-1$ edges 
and a $\Delta$-periodic labeling $\lambda: E \rightarrow \{1,2,\ldots,\Delta\}$ such that, 
for every $i,j$, the duration of the fastest temporal path from $v_i$ to $v_j$ in the $\Delta$-periodic temporal graph $(G,\lambda,\Delta)$ is \emph{at most} $D_{i,j}$.}

\problemdef{\kDeltaLowerBoundLong}
{An $n \times n$ matrix $D$ of positive integers and an integer $k\in \mathbb{N}$.}
{Does there exist a graph $G=(V,E)$ on the vertices $\{v_1,\ldots,v_{n}\}$ with $n+k-1$ edges 
and a $\Delta$-periodic labeling $\lambda: E \rightarrow \{1,2,\ldots,\Delta\}$ such that, 
for every $i,j$, the duration of the fastest temporal path from $v_i$ to $v_j$ in the $\Delta$-periodic temporal graph $(G,\lambda,\Delta)$ is \emph{at least} $D_{i,j}$.}
\todo[inline]{NOT TEMPORALLY INTERESTING: the difficulty is to find a static graph with equal labels}

\todo[inline]{The matrix $D$ is m x m, and capturing the distances from an edge e to an edge f (zoom 14 Nov 22, GH).}
\todo[inline]{Check if the reduction works for the UPPER BOUNDED version (seems to work)}
\todo[inline]{how to extend the poly algorithms of cycles? Easy candidate: FES, FVS, series-parallel graphs, or tw=2 graphs}
\todo[inline]{FVS, $c$-factor approximation, hardness of approximation for $c<1.5$}

@@

\section{
\texorpdfstring{\deltaExact\ } {Simple Delta-TGR}}

In this section we study the hardness of \deltaExact. We first start with defining certain notions, that will be of use when solving the problem.

\todo[inline]{NK:Following definition and observation are copied from the previous section (need to delete from one part).}

    \begin{definition}[Travel delays]
    Let $(G, \lambda, \Delta)$ be a temporal graph satisfying conditions of problem \deltaExact.
    Let $e_1=uv$ and $e_2=vz$ be two incident edges in $G$ with $e_1 \cap e_2 = v$.
    We define the \emph{travel delay} from $u$ to $z$ at vertex $v$, denoted with $\tau_v^{uz}$,
    as the difference of the labels of $e_2$ and $e_1$, where we subtract the value of the label of $e_1$ from the label of $e_2$,  modulo~$\Delta$.
    More specifically:
    \begin{equation}\label{eq:def-VertexWaitingTime}
       \tau_v^{uz} \equiv \lambda (e_2) - \lambda(e_1) \pmod \Delta.
    \end{equation}
    Similarly, $\tau_v^{zu} \equiv \lambda (e_1) - \lambda(e_2) \pmod \Delta$.
    \end{definition}
    Intuitively, the value of $\tau_v^{uz}$ represents how long a temporal path waits at vertex $v$ when first taking the edge $e_1=uv$ and then the edge $e_2 = vz$.

    \begin{observation}\label{obs:travel-delays-both-directions}
    Let $(G, \lambda, \Delta)$ be a temporal graph satisfying conditions of problem \deltaExact.
    For any two incident edges $e_1 = uv$ and $e_2 = vz$ on vertices $u,v,z \in V$, with $e_1 \cap e_2 = v$, we have $\tau_v^{zu} = \Delta - \tau_v^{uz} \pmod \Delta$.
    \end{observation}

@@

\begin{claim} \label{claim:unique-fastest-path-allDelays}
    Let $(G, \lambda, \Delta)$ be a temporal graph satisfying conditions of problem \deltaExact,
    and let $P=P_{1,k}$ be a fastest temporal path from $u=v_1$ to $v=v_k$ on $k$ vertices $v_1,v_2,\dots,v_k$.
    Let us denote with $P_{1,i}$ the sub-path of temporal path $P_{1,k}$, that starts at $v_1$ and finishes at $v_i$.
    Suppose that $P_{1,i}$ is also the fastest temporal path from $u=v_1$ to any other vertex $v_i$ in $P$.
    Then we can determine travel delays on $P$ using the following equation
    \begin{equation}\label{eq:trav-delays-paths}
        \tau_{v_i}^{v_{i-1},v_{i+1}} = D_{1,i+1} - D_{1,i},
    \end{equation}
    for all $i \in \{2,3, \dots, k-1\}$.
\end{claim}

\begin{proof}
    Let $P$ be a fastest temporal path from $v_1$ to $v_k$ with the properties from the claim, and let $v_i$ be an arbitrary vertex in $P \setminus \{v_1,v_k\}$.
    Using the properties of fastest paths and the definition of duration, we can rewrite \cref{eq:trav-delays-paths} as follows
    \begin{align*}
        \tau_{v_i}^{v_{i-1},v_{i+1}} & = D_{1,i+1} - D_{1,i} =  d(P_{1,i+1}) - d(P_{1,i}) \\
        & = 
        \left(\lambda(v_{i}v_{i+1} - \lambda(v_1v_2) + 1\right) -  
        \left(\lambda(v_{i-1}v_{i} - \lambda(v_1v_2) + 1\right) \\
        & = \lambda(v_{i}v_{i+1}) - \lambda(v_{i-1}v_{i}),
        \end{align*}
    which is exactly the definition of $\tau_{v_i}^{v_{i-1},v_{i+1}}$.
\end{proof}

\subsection{
\texorpdfstring{\deltaExact\ } {Simple Delta-TGR}
is polynomial-time solvable on trees and paths}

Let $D$ be a matrix from \deltaExact, when the underlying graph $G$ of $D$ is a tree on $n$ vertices $\{v_1, v_2, \dots, v_n\}$.
Let $v_i,v_j$ be two arbitrary vertices in $G$. Then we know that there exists a unique (static) path $P$ among them.
Consequently, it follows that the temporal paths $P_{i,j}$ from $v_i$ to $v_j$ and $P_{j,i}$ from $v_j$ to $v_i$ are also unique, up to modulo of the period $\Delta$ of the labeling $\lambda$,
and therefore are the fastest.
Then $D$ is of the following form:
\begin{equation*}
    D_{i,j} =
    \begin{cases}
    0 & \text{if $i = j$}, \\
    1 & \text{if $v_i$ and $v_j$ are neighbours in G}\\
    d(P_{i,j}) & \text{else}
    \end{cases},
\end{equation*}
where $d(P_{i,j})$ is the duration of the (unique) temporal path $P_{i,j}$ from $v_i$ to $v_j$.

\begin{observation}\label{obs:travel-delays-in-trees}
    Let $v_i,v_j$  be arbitrary two vertices in a tree $G$. 
    Since there is a unique temporal path $P_{i,j}$ from $v_i$ to $v_j$, it is also the fastest one, therefore $d(P_{i,j}) = D_{i,j}$. 
    Note, all other vertices $v' \in P_{i,j} \setminus \{v_i,v_j\}$ are reached form $v_i$ using a part of the path $P_{i,j}$.
    Now using \cref{claim:unique-fastest-path-allDelays}, we can determine the waiting times for all \emph{inner} vertices of the path $P_{i,j}$.
\end{observation}

\begin{theorem}
    \deltaExact\ can be solved in polynomial time on trees.
\end{theorem}

\begin{proof}
    Let $D$ be an input matrix for problem \deltaExact of dimension $n \times n$.
    Let us fix the vertices of the corresponding graph $G$ of $D$ as $v_1, v_2, \dots, v_n$, where vertex $v_i$ corresponds to the row and column $i$ of matrix $D$.
    This can be done in polynomial time as we need to read through the matrix $D$ once and connect vertices $v_i, v_j$ for which $D_{i,j} = 1$. At the same time we also check if $D_{i,i} = 0$, for all $i \in [n]$.
    When $G$ is constructed we run DFS algorithm on it and check if it has no cycles.
    If at any step we encounter a problem, our algorithm stops and returns a negative answer.

    From now on we can assume that we know that the underlying graph $G$ is a tree and we know how it looks like.
    For the next part of the algorithm we use \cref{obs:travel-delays-in-trees}.

    For a vertex $v_i$ we check which vertex $v_j \in G$ is furthest away from it, \ie for which vertex $v_j$ the entry $D_{i,j}$ is maximum.
    We now take the unique (fastest) temporal path $P_{i,j}$ and using \cref{obs:travel-delays-in-trees} calculate waiting times at all inner vertices.
    We save those values in a matrix $T$ and mark vertices of the path  $P_{i,j}$ as visited.
    Matrix $T$ stores at the position $(i,i-1,i+1)$ the value corresponding to the travel delay at vertex $v_i$ when traveling from $v_{i-1}$ to $v_{i}$, \ie $\tau_{v_i}^{v_{i-1}, v_i}$, whenever $v_{i-1}v_i, v_iv_{i+1} \in E(G)$. All other values of $T$ are set to \textsc{Null}.
    Now we repeat procedure, from vertex $v_i$, for vertices that are not marked as visited yet, \ie vertices in $V \setminus P_{i,j}$.
    We find a vertex in $V \setminus P_{i,j}$ that is furthest away from $v_i$ and repeat the procedure.
    When we have exhausted the $i$-th row of $D$, corresponding to the vertex $v_i$,
    we continue and repeat the procedure for all other vertices.
    If at any point we get two different values for the same travel delay at a specific vertex, then we stop with the algorithm and return the negative answer.
    If the above procedure finishes successfully we get the matrix $T$ with travel delays for all vertices $v_i$ in the tree, of degree at least $2$.
    The above calculation is performed in polynomial time, as for every vertex $v_i$ we inspect the whole graph once.
    \begin{claim}\label{claim:matrixT-travelDelays-tree}
        Matrix $T$ consists of travel delays of all vertices of degree at least $2$ in $G$.
    \end{claim}
    \begin{proof}[Proof of \cref{claim:matrixT-travelDelays-tree}]
        Note, by the definition of travel delays, a vertex of degree $1$ cannot have a travel delay.
        Suppose now that there is a vertex $v_i \in V(G)$ of degree at least $2$, for which our algorithm did not calculate its travel delay.
        Let $v_a, v_b$ be two arbitrary neighbors of $v_i$, \ie $v_a v_i, v_i v_b \in E(G)$.
        Since $G$ is a tree, the unique (and fastest) temporal path from $v_a$ to $v_b$ passes through $v_i$.
        When our algorithm was inspecting the row of $D$ corresponding to the vertex $v_a$, it had to consider the temporal path from $v_a$ to $v_b$. 
        At this point, it calculated $\tau_{v_i}^{v_a,v_b}$. 
        Since this is true for any two $v_a, v_b \in N(v_i)$, it cannot happen that some travel delay at $v_i$ is not calculated.
        Since $v_i$ was an arbitrary vertex in $G$ of degree at least $2$, the claim holds.
    \end{proof}

    Now, given the matrix of travel delays $T$, we can find a labeling $\lambda$ that satisfies $D$.
    We start by fixing a label of one edge as $a$, where $a \in [\Delta]$.
    Knowing the label of the edge $v_1v_2$ and all waiting times in $T$, we can uniquely determine the labels of all other edges.
    More specifically, if we know that $\lambda(v_i v_j) = a$, then for all $v_k \in N(v_i)$ (resp.~$v_{k'} \in N(v_j)$)
    the value $\lambda(v_iv_k) \equiv a + \tau_{v_i}^{v_j,v_k} \pmod \Delta $ (resp.~$\lambda(v_j v_{k'}) \equiv a + \tau_{v_j}^{v_i,v_{k'}} \pmod \Delta $).
    Since there are $\Delta$ options to fix the first label, we can find $\Delta$ different labelings satisfying $D$.
    Note, w.l.o.g we can start determining the labeling $\lambda$ by setting $\lambda(v_1v_2) = 1$.
    It is not hard to see, that also the calculation of the labeling $\lambda$ takes polynomial time, as we have to traverse the graph exactly once, to successfully fix the labeling. Therefore all together the whole algorithm is performed in polynomial time.
\end{proof}

\subsection{
\texorpdfstring{\deltaExact\ } {Simple Delta-TGR}
is polynomial-time solvable on cycles}

Let us observe some properties of the matrix $D$ from \deltaExact, when the underlying graph $G$ of $D$ is a cycle $C_n = \{v_1, v_2, \dots, v_n\}$  on $n$ vertices.
By the definition, each vertex is on distance $0$ from itself. This corresponds with all diagonal entries of $D$ being $0$.
Now, let us observe that each vertex $v_i$ has exactly two neighbours in $C_n$, namely $v_{i-1}$ and $v_{i+1}$,
therefore for all $i\in [n]$ we set $D_{i,i-1} = D_{i, i+1} = 1$, where indices are taken modulo $n$.
This results in the upper and lower diagonal of $D$ having all $1$'s, together with $D_{1,n} = D_{n,1} = 1$.
The matrix $D$ is of the following form
\begin{equation*}
  D = 
  \begin{bmatrix}
    0 & 1 &  &  &  & 1\\
    1 & 0 & 1 &  &  & \\
    & 1 & 0 & 1 & & \\
    & & \ddots & \ddots & \ddots &   \\
    &   & &1 &  0 & 1\\ 
    1 &  &  &  &1 &  0 
  \end{bmatrix},
\end{equation*}
where the empty entries consists of positive integers different than $1$.

Given a matrix in the input of \deltaExact\ we can check in $O(n^2)$ time if it is of correct form, by traversing it once. If it is not of correct for, our algorithm stops and returns the negative answer.
From now on, we will assume, that the input matrix is of correct form.

Let $v_i$ be an arbitrary vertex in the cycle $C_n =\{v_1, v_2, \dots, v_n\}$.
Vertex $v_i$ can reach an arbitrary vertex $v_k \in C_n$ using a positive side of the cycle (\ie going from $v_i$ to $v_{i+1},  v_{i+2}$, etc.), 
and using the negative side of the cycle (\ie going from $v_i$ to $v_{i-1}, v_{i-2}$, etc.).
Let us denote with the $d^+(i,k)$ the duration of the temporal path from $v_i$ to $v_k$ using the positive side of the cycle,
and with $d^-(i,k)$, the duration of the temporal path from $v_i$ to $v_k$ using the negative side of the cycle.
Since these two are the only possible paths from $v_i$ to $v_k$ in $C_n$ we know that $D_{i,k} = min \{d^+(i,k), d^-(i,k)\}$.

\begin{claim} \label{claim:poly-delta-cycle-monotonicity}
    If vertex $v_i$ reaches vertex $v_j$ the fastest, using the positive (resp.~negative) side of the cycle, \ie $v_i, v_{i+1}, \dots, v_{j-1}, v_j$ (resp. $v_i, v_{i-1}, \dots, v_{j+1}, v_j$),
    then $v_i$ reaches all other vertices $v_k$, where $k \in \{i+1, i+2, \dots, j-1\}$ (resp.~ $k \in \{i-1, i-2, \dots, j+1\}$)
    using the same, positive (resp.~negative) side of the cycle,
    where the indices are taken modulo $n$.
\end{claim}
\begin{proof}
    Let $v_i, v_j$ be two arbitrary vertices in $C_n$ and suppose that $D_{i,j} = d^+(i,j)$,
    but there exists a vertex $v_k$ from $C_n$, where $k \in \{i+1, i+2, \dots, j-1\}$,
    for which $D_{i,k} \neq d^+(i,k)$.
    Then $d^-(i,k)$ passes through the vertex $v_j$, and we can split the path from $v_i$ to $v_k$ into two pieces, 
    first from $v_i$ to $v_j$ and the second one from $v_k$ to $v_j$.
    So 
    \begin{equation} \label{eq:poly-delta-cycle-monotonicity}
        d^-(i,k) = d^-(i,j) + d^-(j,k)^*,
    \end{equation} where $d^-(j,k)^*$ is the duration of the path from $v_j$ to $v_k$, using the negative side of the cycle, with consideration that we come to the vertex $v_j$ at time $d^-(i,j)$ and we potentially have to wait there for some positive amount of time, before we continue to $v_j$
    (\ie $d^-(j,k)^* \geq d^-(i,j) \geq D_{j,k}$ includes some waiting time at vertex $v_j$).
    By the assumption we know that $d^+(i,j) \leq d^-(i,j)$, so we can rewrite \cref{eq:poly-delta-cycle-monotonicity} as
    $d^-(i,k) \leq d^+(i,j) + d^-(j,k)^*$. This means that we can reach $v_k$ from $v_j$ by going from $v_i$ to $v_j$ using the positive side of the cycle, wait some time at $v_j$, before we continue back to $v_j$.
    In the above construction vertex $v_k$ is visited twice. By the definition, the fastest temporal path from $v_i$ to $v_k$ visits $v_k$ exactly once. Therefore we can stop at $v_k$ already when traveling from $v_i$ to $v_j$ the first time, using the positive side of the cycle.
    It follows that $D_{i,j} = d^+(i,j) \leq d^-(i,j)$, which contradicts with our assumption.
\end{proof}

Let $v_i$ and $v_j$ be two arbitrary vertices in the cycle $C_n$, for which $v_i v_j \notin E(C_n)$.
Suppose that $P^+_{i,j}$ (resp.~$P^-_{i,j})$ is the fastest temporal path from $v_i$ to $v_j$, using the positive (resp.~negative) side of the cycle, 
\ie $v_i, v_{i+1}, \dots, v_{j-1}, v_j$ (resp. $v_i, v_{i-1}, \dots, v_{j+1}, v_j$).
Then by \cref{claim:poly-delta-cycle-monotonicity} and \cref{claim:unique-fastest-path-allDelays},
we get that we can determine all travel delays at vertices of $P^+_{i,j} \setminus \{v_i, v_j\}$ (resp.~$P^-_{i,j} \setminus \{v_i, v_j\}$).

Let $v_i$ be an arbitrary vertex of $C_n$. Let us look at the row $i$ of the matrix $D$, which corresponds to the durations of fastest temporal paths from $v_i$ to all other vertices $v_j \in C_n$.
Using \cref{claim:poly-delta-cycle-monotonicity} we know that $v_i$ will reach some consecutive vertices $v_{i+1}, v_{i+2}, \dots, v_j$ the fastest, using the positive side of the cycle 
and $v_{i-1}, v_{i-2}, \dots, v_{j+1}$ the fastest, using the negative side of the cycle.
Suppose $v_j \in C_n$ is the last vertex $v_i$ reaches using the positive side of the cycle, and $v_{j+1}$ the last vertex that is reached using the negative side of the cycle.
Then we know that $D_{i,i+1} < D_{i,i+2} < \cdots< D_{i,j-1} < D_{i,j}$ and 
$D_{i,i-1} < D_{i,i-2} < \cdots < D_{i,j+2} < D_{i,j+1}$.
Note that it can happen that $v_j = v_{j+1}$, \ie to reach vertex $v_j$ from $v_i$ the fastest, we can use either positive or negative side of the cycle.
Using the above observations, every row $i$ ($i \in [n]$) of matrix $D$ has two (or one) maximum elements, one at position $j$ and the other at position $j+1$, where $j, j+1 \in [n]$ and the indices are considered modulo $n$.
Let us denote these two values as $m_i^1$ and $m_i^2$.
The row $i$ of $D$ is of the following form, it has a $0$ at the entry $i$, it has $1$ at entries $i-1, i+1$,
the values increase on the positions $i+1, i+2, i+3, \dots, j-1, j$ for some $j$ with value $m_i^1$, 
and on the other side, values increase on the positions $i-1, i-2, i-3, \dots, j+2, j+1$ for some $j+1$ with value $m_i^2$,
where indices are taken modulo $n$.

Knowing this, we can split the vertices $v_j \in V(C_n) \setminus \{v_i\}$ into two parts, ones that are reached from $v_i$ the fastest using the positive side of the cycle and ones that are reached using the negative side of the cycle.
To determine these two sets we do the following.
We fix a vertex $v_i \in C_n$ and check its corresponding row in the matrix $D$. 
We determine two max values $m_i^1$ and $m_i^2$ at positions $j$ and $j+1$ (modulo $n$), respectively,
for which it has to hold that
$D_{i,i+1} < D_{i,i+2} < \cdots< D_{i,j-1} < D_{i,j}=m_i^1$ and 
$D_{i,i-1} < D_{i,i-2} < \cdots < D_{i,j+2} < D_{i,j+1}=m_i^2$.
Note, it can also happen that $m_i^1 = m_i^2$. 
Now denote the path that uses the positive side of the cycle, from $v_i$ to $v_j$, as $P^+_i$ and
the path that uses the negative side of the cycle, from $v_i$ to $v_{j+1}$, as $P^-_i$.
By \cref{claim:poly-delta-cycle-monotonicity} and \cref{claim:unique-fastest-path-allDelays} we can calculate travel delays at every vertex $v_k \in C_n \setminus \{v_i, v_j, v_{j+1}\}$,
which we store in a list $T$ of length $n$,
where the entry at the position $k$ corresponds to the travel delay at vertex $v_k$ when traveling from $v_{k-1}$ to $v_{k+1}$.
Note, from \cref{obs:travel-delays-both-directions} it follows, that it is enough to store the value of the travel delay in one direction (\ie knowing $\tau_v^{u,w}$ we know also $\tau_v^{w,u}$).
We repeat the above procedure for all rows in the matrix $D$, \ie for all vertices $v_i \in C_n$.
Calculation in one row is performed in $O(n)$ time, repeating it for all rows we need $O(n^2)$ time.

To determine the labeling $\lambda$ satisfying the matrix $D$, we have to make sure that we have calculated travel delays for all vertices.

\begin{claim}
List $T$ of travel delays has non-empty values at all positions, \ie we have successfully calculated travel delays for all vertices.
\end{claim}
\begin{proof}
    Throughout the proof we
    denote with $d^+(i,j)$ the duration of the temporal path that starts at $v_i$ and finishes at $v_j$, that uses the positive side of the cycle $C_n$,
    and similarly with $d^-(i,j)$ the duration of the temporal path that starts at $v_i$ and finishes at $v_j$, that uses the negative side of the cycle $C_n$,
    where indices are taken modulo $n$.

    Suppose for the contradiction that the statement of the claim does not hold. Then there exists a vertex $v_i \in C_n$, for which we did not calculate its travel delay.
    Let $v_j \in C_n \setminus \{v_i\}$ be an arbitrary vertex. 
    Note that the only time we cannot calculate the travel delay at vertex $v_i$, when considering vertex $v_j$, is in this case when $v_i$ is one of the maximum elements in $D_j$, \ie $v_i$ is a vertex that is on a maximum duration from $v_j$.
    This has to hold for all vertices $v_j$, therefore $v_i$ has to be on the maximum distance from every vertex $v_j \in C_n \setminus \{v_i\}$.
    It then also has to hold for vertices $v_{i-1}, v_{i+1}$, that are neighbours of $v_i$.
    We know that in this case $d^+(i-1,i) = 1, d^-(i+1,i) = 1$.
    Since $v_i$ has to be on the maximum duration from both of them 
    (\ie is one of the two maximum values $m^1_{i-1}, m^2_{i-1}$ for vertex $v_{i-1}$ and one of the two maximum values $m^1_{i+1}, m^2_{i+1}$ for vertex $v_{i+1}$),
    we know that 
    \begin{align}\label{eq:cycle-travel-delays-all}
        & d^+(i+1,i-1) < d^-(i+1,i-1) & \text{and}\\
        & d^-(i-1,i+1) < d^+(i-1,i+1).
    \end{align}
    If this would not hold, the fastest path would go through $v_i$ and we would be able to calculate the travel delay at $v_i$.

    Denote the labels $\lambda(v_{i-1}v_i) = a, \lambda(v_{i}v_{i+1})=a'$, and $\lambda(v_{i-2}v_{i-1} = b, \lambda (v_{i+1}v_{i+2})=b'$, where $a,a',b,b' \in [\Delta]$.
    W.l.o.g. we can suppose that $a \geq a'$.
    Therefore, using the definition for the duration of temporal paths, we get that
    $d^+(i+1,i-1) = (k \delta + b) - b' + 1$, where $k$ is some non-negative integer, and $d^-(i+1,i-1) = a - a' + 1$.
    Using the first inequality from \cref{eq:cycle-travel-delays-all} we get that 
    $(k \delta + b) - b' +1 < a - a' + 1$ which is equivalent to $(k \delta + b) - b' < a - a'$, which can be true only when $k = 0$, but because the duration is positive we get that $b > b'$ and
    $b - b' < a - a'$.
    Now again using the definition of the duration, we get
    $d^+(i-1,i+1)= (k\delta + a') - a + 1$, where $k$ is some non-negative integer, but since $v_{i-1}v_i$ and $v_iv_{i+1}$ are incident edges, we know also that $k=1$, therefore $d^+(i-1,i+1)= \delta + a - a + 1$. 
    Again by the definition, $d^-(i-1,i+1) = (k' \delta + b') -b + 1$, for some non-negative integer $k'$.
    Using the second inequality from \cref{eq:cycle-travel-delays-all} we get that 
    $(k' \delta + b') -b + 1 < \delta + a - a + 1$.
    Which is equivalent to  $(k' \delta + b') -b < \delta + a' - a $. This can hold only when $k' = 1$, in that case we get $ \delta + b' -b < \delta + a' - a $ which is equivalent to $b' -b < a' - a$.
    This is in the contradiction with the inequality  $b - b' < a - a'$ as this is equivalent to $b'-b > a'-a$.

    Therefore it cannot happen that there is a vertex $v_i$ for which we cannot calculate its travel delay in $C_n$.
\end{proof}

All of the above observations imply the following result.
\begin{theorem}
    \deltaExact\ can be solved in polynomial time on cycles.
\end{theorem}
